A Novel Solution to Resonant Scattering with Self-Consistent Boundary Conditions

### 2022-04-06 ms.tex
Addressing reviewer comments up to the appendix

In Figure \ref{fig:tau_scaling}, the late-time decay timescale of the wait time distribution is shown as a function of $\tau_0$. It is shown that the time constant of exponential decay in fitted Monte Carlo escape time distributions converges with $\gamma_{11}^{-1}$ at sufficiently high $\tau_0$, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower $\tau_0$, the effects of line core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately includes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile, and also does not use enough spatial modes to accurately model the frequency regime near line center. However, as $\tau_0$ grows, the effect of core scattering becomes smaller and the approximations hold, agreeing well with the expected $(a\tau_0)^{1/3}$ scaling \citep{1975ApJ...201..350A} for the rate constant at late times.

@@

In Figure \ref{fig:steadystate}, the fluence for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated''), and is compared with two analytic solutions: the steady-state $H_0$ solution (Equation \ref{eq:H0surf}) with no time dependence (labelled ``Steady State''), and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in the first line of Equation (\ref{eq:H0surf}). Additional spatial modes $n$ increase the solutions' accuracy in the core of the line. If more spatial modes are included, the agreement with the steady-state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x=x_s=0$. If additional frequency modes are included, faster-oscillating terms are incorporated into the Fourier sum over eigenmodes which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would primarily affect the distribution at early times since the late time distribution is determined by the lowest order modes. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated.

\begin{figure}
    \centering
    \includegraphics{steadystate.pdf}
    \caption{Fluence $P(x)$ vs. $x$. The steady-state and time-integrated spectra for $n=1, ..., 20$ and $m=1, ..., 500$ are shown with $x_s=0$ and $\tau_0=10^7$.}
    \label{fig:steadystate}
\end{figure}

@@

\subsection{Steady-State Source}
A primary goal of this work is to present a solution for resonant scattering of photons near the line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \citet{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We introduce a new term to this solution, $J_{\rm bc}$, which allows the boundary condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This is solved using a continuous Fourier expansion in frequency. The integrals are discretized and the Fourier coefficients solved for numerically. The resulting flux correction, $H_{\rm bc}$, scales as $H_0(a\tau_0)^{-1/3}$. Thus, for large $a\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $a\tau_0$. Since the Laplacian form of the differential equation is only correct when the line profile is $\phi \approx a/(\pi x^2 \Delta)$, our solutions do not accurately model the Doppler core of the \lya line. Because the peak of the spectral energy distribution of escaping photons is $x_{\rm peak} {\sim} (a\tau_0)^{1/3}$, calculations performed at small $a\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions for $a\tau_0 \geq 1$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition presented in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, and \citet{2006ApJ...649...14D}. Several papers have previously compared these analytic models to Monte Carlo and seen discrepancies on the order of this correction. For example, in the top-left panel of Figure 1 from \citet{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum ($H_0$) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Another example is in \citet{2015MNRAS.449.4336S}, where the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent. Both of these solutions are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$. We show in our Figure \ref{fig:sol_mc_tau} that the error present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$ for T$=10^4$ K, corresponding to $a=4.72 \times 10^{-4}$. We note that our corrective term $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in the aforementioned solutions.

@@

There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}, but these methods are not sufficient for our application since they do not track the amount of time photons spend in the line core, which is needed to tabulate the intensity. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}. In this approach, an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \citet{1990ApJ...350..216N} solution, but this solution of course does not utilize the frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more.

>>>>

In Figure \ref{fig:tau_scaling}, the late-time decay timescale of the wait time distribution is shown as a function of $\tau_0$. It is shown that the time constant of exponential decay in fitted Monte Carlo escape time distributions converges with $\gamma_{11}^{-1}$ at sufficiently high $\tau_0$, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower $\tau_0$, the effects of line core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately includes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile, and also does not use enough spatial modes to accurately model the frequency regime near line center. However, as $\tau_0$ grows, the effect of core scattering becomes smaller and the approximations hold, agreeing better with the expected $(a\tau_0)^{1/3}$ scaling \textbf{\citep{1975ApJ...201..350A}} for the rate constant at late times. \textbf{The excess in the Monte Carlo data points due to core scattering decreases exponentially at higher $\tau_0$, and though these points are not shown at $\tau_0=10^8$ and $10^9$ due to computational expense, it is expected that the fractional error between the Monte Carlo and the $(a\tau_0)^{1/3}$ scaling would be less than 2\% at $\tau_0=10^9$.}

@@

In Figure \ref{fig:steadystate}, the fluence for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated''), and is compared with two analytic solutions: the steady-state $H_0$ solution (Equation \ref{eq:H0surf}) with no time dependence (labelled ``Steady State''), and the result for summing \textbf{a finite number of} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in the first line of Equation (\ref{eq:H0surf}). Additional spatial modes $n$ increase the solutions' accuracy in the core of the line. If more spatial modes are included, the agreement with the steady-state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x=x_s=0$. If additional frequency modes are included, faster-oscillating terms are incorporated into the Fourier sum over eigenmodes which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would primarily affect the distribution at early times since the late time distribution is determined by the lowest order modes. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated.

\begin{figure}
    \centering
    \includegraphics{steadystate.pdf}
    \caption{Fluence $P(x)$ vs. $x$ \textbf{(see Equation \ref{eq:spectrum}). Fluence is the radiation flux integrated over time. }Steady-state and time-integrated spectra for $n=1, ..., 20$ and $m=1, ..., 500$ are shown with $x_s=0$ and $\tau_0=10^7$. \textbf{Note that the x-axis begins near the line core, as we are only concerned with the solutions' accuracy in the line wing.}}
    \label{fig:steadystate}
\end{figure}

@@

\subsection{Steady-State Source}
A primary goal of this work is to present a solution for resonant scattering of photons near the line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \citet{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We introduce a new term to this solution, $J_{\rm bc}$, which allows the boundary condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This is solved using a continuous Fourier expansion in frequency. The integrals are discretized and the Fourier coefficients solved for numerically. The resulting flux correction, $H_{\rm bc}$, scales as $H_0(a\tau_0)^{-1/3}$. Thus, for large $a\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $a\tau_0$. Since the Laplacian form \textbf{for frequency redistribution in }the differential equation is only correct \textbf{for photons in the wing where the} line profile is $\phi \approx a/(\pi x^2 \Delta)$, our solutions do not accurately model the Doppler core of the \lya line. Because the peak of the spectral energy distribution of escaping photons is $x_{\rm peak} {\sim} (a\tau_0)^{1/3}$, calculations performed at small $a\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions for $a\tau_0 \gg 1$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition presented in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, and \citet{2006ApJ...649...14D}. Several papers have previously compared these analytic models to Monte Carlo and seen discrepancies on the order of this correction. For example, in the top-left panel of Figure 1 from \citet{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum ($H_0$) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Another example is in \citet{2015MNRAS.449.4336S}, where the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent. Both of these solutions are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$. We show in our Figure \ref{fig:sol_mc_tau} that the error present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$ for T$=10^4$ K, corresponding to $a=4.72 \times 10^{-4}$. We note that our corrective term $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in the aforementioned solutions.

@@

There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}, but these methods are not sufficient for our application since they do not \textbf{directly measure or sample} the amount of time photons spend in the line core, which is needed to tabulate the intensity. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}. In this approach, an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \citet{1990ApJ...350..216N} solution, but this solution of course does not utilize the frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more.
